---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Alhaylam Alajmi"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---


## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.



**In this analysis, I examine median home values across counties in Texas and ask which counties are measured most and least reliably. I chose this topic because my experience in major infrastructure and urban development projects has increased my interest in understanding how development outcomes differ across communities.**

---

## Choosing what to measure



```{r}
my_state <- "TX"        # two-letter abbreviation, in quotes
my_variable <- "B25077_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


The dataset contains 254 rows, which matches the 254 counties in Texas

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```



Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```


Presidio County has the largest margin of error relative to its estimate, at 63.9%, making it the least reliably measured county in this analysis.

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```



Of the 254 Texas counties, 53 are classified as high confidence 86 as moderate, and 115 as low confidence.

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```



High confidence counties have an average population of 482,559, compared with only 6,426 in low confidence counties. Moderate confidence counties have an average population of 28,971. This pattern suggest that estimates are less reliable in smaller counties.

---

## What this means for policy



The results show that some small counties in Texas have less reliable estimates of median home values than larger counties. Presidio County has the highest relative margin of error at 63.9%. It is followed by Motley County at 59.5% and Kent County at 56.5%. Crockett, Childress, and Sterling counties also have high margins of error. This means we should be careful when using the estimates for these counties because the real values may be different from the published estimates.

This pattern is connected to the way the American Community Survey works. The ACS does not count every person or home. Instead, it uses a sample to estimate values for the whole community. Smaller counties usually have fewer people and homes in the sample, so their estimates have more uncertainty. In my results, high-confidence counties have an average population of 482,559, while low-confidence counties have an average population of only 6,426. Moderate-confidence counties have an average population of 28,971. This shows a clear relationship between smaller population and lower reliability.

This uncertainty can cause problems if a state agency uses median home value to distribute funding and assumes every county estimate is equally accurate. If a small county’s home value is overestimated, it may appear wealthier and receive less funding than it needs. Rural and small communities may therefore be affected the most. Agencies should consider both the estimate and its margin of error before making funding decisions.


---



